---
title: "ChIP-seq assignment"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(GenomicRanges)
library(rtracklayer)
library(ggplot2)
library(dplyr)
```

# Overview

This report processes Drosophila CTCF ChIP-seq data from raw FASTQ and reports:

1. number of mapped reads and mapping percentage  
2. number of peaks found  
3. one plot showing the signal around a peak located inside a gene  

---

# Input files

```{r}
fastq_file   <- "ENCFF127RRR.fastq.gz"
bam_file     <- "ctcf.sorted.bam"
flagstat_file <- "ctcf.flagstat.txt"
peak_file    <- "ctcf_peaks.narrowPeak"
gtf_file     <- "dmel-all-r6.57.gtf"   # replace with your own annotation file
bedgraph_file <- "ctcf.signal.bedGraph" # optional, for plotting if available

fastq_file
bam_file
flagstat_file
peak_file
gtf_file
```

# Processing commands

The raw FASTQ was processed externally with a standard ChIP-seq workflow. Example commands are shown below.

## 1. Download FASTQ

```{bash eval=FALSE}
curl -L -o ENCFF127RRR.fastq.gz \
https://www.encodeproject.org/files/ENCFF127RRR/@@download/ENCFF127RRR.fastq.gz
```

## 2. Align reads to Drosophila genome

```{bash eval=FALSE}
curl -O https://hgdownload.soe.ucsc.edu/goldenPath/dm6/bigZips/dm6.fa.gz
gunzip dm6.fa.gz
bowtie2-build dm6.fa dm6_index
bowtie2 -x dm6_index -U ENCFF127RRR.fastq.gz -S ctcf.sam
samtools view -bS ctcf.sam > ctcf.bam
samtools sort ctcf.bam -o ctcf.sorted.bam
samtools index ctcf.sorted.bam
samtools flagstat ctcf.sorted.bam > ctcf.flagstat.txt
```

## 3. Peak calling

```{bash}
macs2 callpeak \
-t ctcf.sorted.bam \
-f BAM \
-g dm \
-n ctcf
```

## 4. Optional signal track for plotting

```{bash eval=FALSE}
bedtools genomecov -ibam ctcf.sorted.bam -bg > ctcf.signal.bedGraph
```

---

# 1. Mapping statistics

```{r}
flagstat_lines <- readLines(flagstat_file)

mapped_line <- flagstat_lines[grepl(" mapped \\(", flagstat_lines)][1]
mapped_line
```

```{r}
# Extract mapped reads and percentage from samtools flagstat output
mapped_reads <- sub("^([0-9]+) \\+ [0-9]+ mapped .*", "\\1", mapped_line)
mapped_pct   <- sub("^.*\\(([0-9.]+)%.*\\)$", "\\1", mapped_line)

mapped_reads <- as.numeric(mapped_reads)
mapped_pct   <- as.numeric(mapped_pct)

mapped_reads
mapped_pct
```

**Mapped reads:** `r format(mapped_reads, big.mark=",")`  
**Mapping percentage:** `r mapped_pct`%

---

# 2. Number of peaks found

```{r}
peaks <- read.table(peak_file, sep = "\t", header = FALSE, stringsAsFactors = FALSE)
n_peaks <- nrow(peaks)
n_peaks
```

**Number of peaks found:** `r format(n_peaks, big.mark=",")`

---

# 3. Find a peak located inside a gene

The narrowPeak format has at least the first 3 genomic columns: chromosome, start, end.

```{r}
peak_gr <- GRanges(
  seqnames = peaks[[1]],
  ranges = IRanges(start = peaks[[2]] + 1, end = peaks[[3]])
)
```

```{r}
library(rtracklayer)
library(GenomicRanges)

gtf <- import("../dm6.refGene.gtf")
peak_gr <- rtracklayer::import("ctcf_peaks.narrowPeak", format = "narrowPeak")

genes_gr <- gtf[mcols(gtf)$type == "transcript"]

hits <- findOverlaps(peak_gr, genes_gr)

if (length(hits) == 0) stop("No peak overlapping transcript regions found.")

peak_in_gene_idx <- queryHits(hits)[1]
gene_idx <- subjectHits(hits)[1]

selected_peak <- peak_gr[peak_in_gene_idx]
selected_gene <- genes_gr[gene_idx]

selected_peak
selected_gene
```

```{r}
gene_name <- NA
if ("gene_name" %in% names(mcols(selected_gene))) {
  gene_name <- mcols(selected_gene)$gene_name
} else if ("gene_id" %in% names(mcols(selected_gene))) {
  gene_name <- mcols(selected_gene)$gene_id
}
gene_name
```

Chosen peak is inside gene: `r as.character(gene_name)`

---

# 4. Plot signal around the selected intragenic peak

This section uses a bedGraph signal track if available.

```{r}
signal <- read.table(
  bedgraph_file,
  sep = "\t",
  header = FALSE,
  stringsAsFactors = FALSE,
  col.names = c("chr", "start", "end", "score")
)

head(signal)
```

```{r}
peak_chr <- as.character(seqnames(selected_peak))
peak_center <- round((start(selected_peak) + end(selected_peak)) / 2)

window_start <- peak_center - 1000
window_end   <- peak_center + 1000

plot_df <- signal %>%
  filter(chr == peak_chr,
         end >= window_start,
         start <= window_end) %>%
  mutate(mid = (start + end) / 2,
         rel_pos = mid - peak_center)

head(plot_df)
```

```{r fig.width=8, fig.height=4}
ggplot(plot_df, aes(x = rel_pos, y = score)) +
  geom_line() +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = paste("CTCF signal around intragenic peak in", gene_name),
    x = "Position relative to peak center (bp)",
    y = "Signal"
  ) +
  theme_bw()
```

---

# Final report summary

```{r}
summary_df <- data.frame(
  metric = c("Mapped reads", "Mapping percentage", "Peaks found"),
  value = c(
    format(mapped_reads, big.mark = ","),
    paste0(mapped_pct, "%"),
    format(n_peaks, big.mark = ",")
  )
)

summary_df
```

**Mapped reads:** `r format(mapped_reads, big.mark=",")`  
**Mapping percentage:** `r mapped_pct`%  
**Peaks found:** `r format(n_peaks, big.mark=",")`  

The figure above shows the signal around one peak located inside a gene.