---
title: "assignment"
author: "Lulu"
date: "2026-05-13"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(rtracklayer)
library(GenomicRanges)
library(pheatmap)
library(ggplot2)
library(dplyr)
library(clusterProfiler)
library(org.Hs.eg.db)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(ChIPseeker)
library(tidyr)
```

```{r download-data}
# 1. Download and unzip only if the folder doesn't exist
if (!dir.exists("clustering_assignment")) {
  download.file(
    url = "https://ethz-ins.org/content/clustering_assignment.zip",
    destfile = "clustering_assignment.zip",
    mode = "wb"
  )
  unzip("clustering_assignment.zip", exdir = "clustering_assignment")
}
```

```{r read-peaks}
# 4. Check files
# R Markdown ignores setwd() across chunks, so we specify the folder directly
peak_files <- list.files(path = "clustering_assignment", pattern = "\\.bed$", full.names = TRUE)

peaks <- lapply(peak_files, rtracklayer::import)

# Keep high-confidence peaks
peaks_high <- lapply(peaks, function(x) {
  x[x$score > 800]
})

# Create union of non-redundant regions
regions <- reduce(unlist(GRangesList(peaks_high)))
```

```{r extract-signals}
# Find bigWig files inside the subfolder
bw_files <- list.files(
  path = "clustering_assignment",
  pattern = "\\.bw$|\\.bigWig$|\\.bigwig$",
  full.names = TRUE,
  recursive = TRUE,
  ignore.case = TRUE
)

# Function to calculate average signal for each region
get_bw_signal <- function(bw_file, regions) {
  bw_signal <- rtracklayer::import(bw_file, which = regions, as = "NumericList")
  signal <- sapply(bw_signal, function(x) {
    if (length(x) == 0) {
      return(0)
    } else {
      return(mean(x, na.rm = TRUE))
    }
  })
  return(signal)
}

# Extract signal from all bigWig files
signal_list <- lapply(bw_files, get_bw_signal, regions = regions)

# Convert list to matrix
signal_matrix <- do.call(cbind, signal_list)

# Add column names
colnames(signal_matrix) <- gsub(
  "\\.bw$|\\.bigWig$|\\.bigwig$",
  "",
  basename(bw_files),
  ignore.case = TRUE
)

# Replace NA with 0
signal_matrix[is.na(signal_matrix)] <- 0
```

```{r filter-and-cluster}
# Make numeric matrix
signal_matrix <- as.data.frame(signal_matrix)

signal_matrix[] <- lapply(signal_matrix, function(x) {
  as.numeric(as.character(x))
})

signal_matrix <- as.matrix(signal_matrix)
signal_matrix[is.na(signal_matrix)] <- 0

# Keep only regions with signal
keep_regions <- rowSums(signal_matrix) > 0

if (sum(keep_regions) >= 2) {
  
  signal_matrix_filtered <- signal_matrix[keep_regions, , drop = FALSE]
  regions_filtered <- regions[keep_regions]
  
  # Scale signal matrix
  signal_scaled <- scale(signal_matrix_filtered)
  
  # Replace problematic values
  signal_scaled[is.na(signal_scaled)] <- 0
  signal_scaled[is.nan(signal_scaled)] <- 0
  signal_scaled[is.infinite(signal_scaled)] <- 0
  
  # Check number of distinct signal patterns
  n_distinct_patterns <- nrow(unique(signal_scaled))
  
  # Choose k safely
  k <- min(4, n_distinct_patterns, nrow(signal_scaled))
  
  if (k >= 1) {
    set.seed(123)
    
    clusters <- kmeans(
      signal_scaled,
      centers = k,
      nstart = 25
    )
    
    regions_filtered$cluster <- clusters$cluster
  }
  
} else {
  message("Not enough non-zero signal regions for k-means clustering.")
  
  # Use original matrix for heatmap instead
  signal_scaled <- scale(signal_matrix)
  signal_scaled[is.na(signal_scaled)] <- 0
  signal_scaled[is.nan(signal_scaled)] <- 0
  signal_scaled[is.infinite(signal_scaled)] <- 0
  
  regions_filtered <- regions
}
```

```{r plot-heatmap}
# pheatmap requires explicit rownames to map the annotation_row correctly
rownames(signal_scaled) <- paste0("Region_", 1:nrow(signal_scaled))

annotation_row <- data.frame(
  Cluster = factor(clusters$cluster)
)
rownames(annotation_row) <- rownames(signal_scaled)

pheatmap(
  signal_scaled,
  annotation_row = annotation_row,
  cluster_rows = TRUE,
  cluster_cols = TRUE,
  show_rownames = FALSE,
  main = "Clustering of CREB-family transcription factor binding"
)
```

```{r plot-average-signal}
# Use filtered matrix and filtered regions to ensure dimensions match perfectly
plot_df <- as.data.frame(signal_matrix_filtered)
plot_df$cluster <- factor(regions_filtered$cluster)

avg_df <- plot_df %>%
  group_by(cluster) %>%
  summarise(across(everything(), mean))

avg_long <- pivot_longer(
  avg_df,
  cols = -cluster,
  names_to = "Factor",
  values_to = "Average_signal"
)

ggplot(avg_long, aes(x = Factor, y = Average_signal, group = cluster, color = cluster)) +
  geom_line() +
  geom_point(size = 2) +
  theme_bw() +
  labs(
    title = "Average binding signal across clusters",
    x = "Transcription factor",
    y = "Average bigWig signal"
  )
```

```{r go-enrichment, message=FALSE, warning=FALSE}
# Select one cluster, for example cluster 1
cluster_regions <- regions_filtered[regions_filtered$cluster == 1]

# Annotate regions to nearest genes
txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene

peak_anno <- annotatePeak(
  cluster_regions,
  TxDb = txdb,
  tssRegion = c(-3000, 3000),
  annoDb = "org.Hs.eg.db"
)

anno_df <- as.data.frame(peak_anno)

genes <- unique(anno_df$geneId)
genes <- genes[!is.na(genes)]

ego <- enrichGO(
  gene          = genes,
  OrgDb         = org.Hs.eg.db,
  keyType       = "ENTREZID",
  ont           = "BP",
  pAdjustMethod = "BH",
  qvalueCutoff  = 0.05,
  readable      = TRUE
)

dotplot(ego, showCategory = 10) +
  ggtitle("GO enrichment for cluster 1")
```

## Discussion and Conclusion

Using high-confidence peak regions on chromosome 1, I compared the binding profiles of three CREB-family transcription factors: CREB1, CREB3, and CREB3L1. The heatmap revealed both shared and distinct binding patterns across the selected genomic regions. Hierarchical clustering suggested that CREB1 and CREB3L1 have more similar binding profiles, whereas CREB3 showed a more distinct pattern.

K-means clustering separated the genomic regions into four clusters. The average signal plot showed that each cluster was characterized by different transcription factor binding patterns. Cluster 2 showed relatively strong CREB1 signal, cluster 3 was dominated by CREB3 signal, and cluster 4 showed strong CREB3L1 signal. This suggests that although CREB-family transcription factors share part of their chromatin binding landscape, some genomic regions display factor-specific binding preferences.

GO enrichment analysis of cluster 1 showed enrichment for biological processes related to RNA splicing, ribonucleoprotein complex biogenesis, protein-RNA complex assembly, spliceosomal complex assembly, and mRNA splice site recognition. These results suggest that genes associated with cluster 1 may be involved in RNA processing and post-transcriptional regulation. Overall, the analysis indicates that CREB-family transcription factors show both shared and differential binding patterns, and that at least one cluster is linked to RNA-processing-related functions.